Converting LaTeX Legacy R Journal Articles into R Markdown Articles using texor and rebib

### 2023-08-23 texor.Rmd
start on the check list of changes

---
title: "Converting LaTeX Legacy R Journal Articles into R Markdown Articles using texor and rebib"
date: "2022-11-09"
abstract: >
  In 2021 the R Journal made a change of templates for article writing to R Markdown instead of LaTeX. The reasons were to encourage better reproducibility of articles using dynamic documents, enable interactivity in articles, and to make the articles more accessible for vision-impaired readers. A resulting challenge was to explore whether legacy articles might be suitably converted into HTML output. This paper describes the process to convert an R Journal article from LaTeX to HTML format via R Markdown, and the two new R packages, \CRANpkg{texor} and \CRANpkg{rebib}, that can be used to achieve the conversion.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Abhishek Ulayil
    affiliation: Institute of Actuaries of India (student)
    address:
    - Mumbai, India
    orcid: 0009-0000-6935-8690
    email: perricoq@outlook.com
  - name: Mitch O'Hara-Wild
    affiliation: Monash University
    address:
    - Melbourne, Australia
    orcid: 0000-0001-6729-7695
    email: mail@mitchelloharawild.com
  - name: Christophe Dervieux
    affiliation: Posit PBC
    address:
    - Paris, France
    orcid: 0000-0003-4474-2498
    email: christophe.dervieux@gmail.com
  - name: Heather Turner
    affiliation: University of Warwick
    address:
    - Coventry, United Kingdom
    orcid: 0000-0002-1256-3375
    email: ht@heatherturner.net
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Melbourne, Australia
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
type: package
output: 
  rjtools::rjournal_web_article:
    web_only: no
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

@@

## Logging the conversion

XXX what should go here?

@@

The package `texor` can be installed from CRAN using:

```r
install.packages("texor")
```

and the development version from https://github.com/Abhi-1U/texor. The website for the package, https://abhi-1u.github.io/texor, has vignettes documenting usage.

@@

XXX - Should we include various options/params for the function here ?

@@

The package `rebib` can be installed from CRAN using:

```r
install.packages("rebib")
```

and the development version from https://github.com/Abhi-1U/rebib. The website for the package, https://abhi-1u.github.io/rebib, has vignettes documenting usage.

@@

No shift to quarto yet.

Known problems that need manual fixes.

>>>>

---
title: "Converting LaTeX Legacy R Journal Articles into R Markdown Articles using texor and rebib"
date: "2022-11-09"
abstract: >
  In 2021 the R Journal made a change of templates for article writing to R Markdown instead of LaTeX. The reasons were to encourage better reproducibility of articles using dynamic documents, enable interactivity in articles, and to make the articles more accessible for vision-impaired readers. A resulting challenge was to explore whether legacy articles might be suitably converted into HTML output. This paper describes the process to convert an R Journal article from LaTeX to HTML format via R Markdown, and the two new R packages, \CRANpkg{texor} and \CRANpkg{rebib}, that can be used to achieve the conversion.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Abhishek Ulayil
    affiliation: Institute of Actuaries of India (student)
    address:
    - Mumbai, India
    orcid: 0009-0000-6935-8690
    email: perricoq@outlook.com
  - name: Mitch O'Hara-Wild
    affiliation: Monash University
    address:
    - Melbourne, Australia
    orcid: 0000-0001-6729-7695
    email: mail@mitchelloharawild.com
  - name: Christophe Dervieux
    affiliation: Posit PBC
    address:
    - Paris, France
    orcid: 0000-0003-4474-2498
    email: christophe.dervieux@gmail.com
  - name: Heather Turner
    affiliation: University of Warwick
    address:
    - Coventry, United Kingdom
    orcid: 0000-0002-1256-3375
    email: ht@heatherturner.net
  - name: Dianne Cook
    affiliation: Monash University
    address:
    - Melbourne, Australia
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib

@@

@@

The package `texor` can be installed from CRAN and the development version from https://github.com/Abhi-1U/texor. The website for the package, https://abhi-1u.github.io/texor, has vignettes documenting usage.

@@

@@

The package `rebib` can be installed from CRAN and the development version from https://github.com/Abhi-1U/rebib. The website for the package, https://abhi-1u.github.io/rebib, has vignettes documenting usage.

# The process of converting all the legacy articles

# Known problems needing manual fixes

@@

Since the shift to R Markdown, a new development for the open source community has emerged: quarto [@quarto]. Although R Markdown would suggest a focus on R, it was always possible to include code chunks written in other languages. But Quarto makes this cleaner, and thus more appealing for non-R developers. It also provides a cleaner type-setting. At some point, the R Journal will likely shift to a Quarto template, which is reasonably straightforward, but for the present R Markdown is a suitable dynamic document delivery system for the R Journal.

### 2023-08-23 texor.tex
start on the check list of changes

The numbering of equations is a bit trickier. LaTeX automatically numbers equations, unless specifically instructed not to. Equation numbering in R Markdown requires specific labeling using \texttt{(\textbackslash{}\#eq:xx)} as described in Yihui Xie (2023). The \CRANpkg{texor} helps by adding the labeling using a Lua filter to convert the existing \texttt{\textbackslash{}label\{..\}} to \texttt{(\textbackslash{}\#eq:xx)}.

@@

\hypertarget{logging-the-conversion}{%
\subsection{Logging the conversion}\label{logging-the-conversion}}

XXX what should go here?

@@

The package \texttt{texor} can be installed from CRAN using:

\begin{verbatim}
install.packages("texor")
\end{verbatim}

and the development version from \url{https://github.com/Abhi-1U/texor}. The website for the package, \url{https://abhi-1u.github.io/texor}, has vignettes documenting usage.

@@

The package \texttt{rebib} can be installed from CRAN using:

\begin{verbatim}
install.packages("rebib")
\end{verbatim}

and the development version from \url{https://github.com/Abhi-1U/rebib}. The website for the package, \url{https://abhi-1u.github.io/rebib}, has vignettes documenting usage.

@@

No shift to quarto yet.

Known problems that need manual fixes.

@@

\begin{itemize}
\tightlist
\item
  \texttt{code-env}: Explains how different code environments defined by the R Journal style are handled, and additional details such as code in figure environments, and code in table environments.
\item
  \texttt{math-env}: Examples of inline math, display math, how equation numbering is handled by a Lua filter to convert from LaTeX labeling to R Markdown labeling.
\item
  \texttt{figure-env}: Explains how the variety of figure definitions are handled in the conversion, including different image formats, numbering, captions, labeling, multiple images, and \texttt{tikz} (Josh Cassidy 2013) images.
\item
  \texttt{table-env}: Examples of how a variety of table types are converted, including multicolumn, complex and wide tables.
\item
  \texttt{lua-filters}: Overview and lots of small examples of Lua filters to handle the custom output needed for the R Markdown format.
\item
  \texttt{metadata}: This has a collection of additional format handling including extracting metadata like author names and affiliations, article identifiers used in the review process, and handling citations, footnotes and links.
\item
  \texttt{bibliography}: The bibliography was handled differently over the years of the journal, and this details how to use the \texttt{rebib} functionality to handle \texttt{bbl} files, embedded \texttt{bbl}, to convert into the standard \texttt{.bib} format.
\end{itemize}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-tableHTML}{}}%
Boutaris, Theo, Clemens Zauchner, and Dana Jomar. 2023. \emph{tableHTML: A Tool to Create HTML Tables}. \url{https://CRAN.R-project.org/package=tableHTML}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-casflow}{}}%
Josh Cassidy. 2013. \emph{{LaTeX Graphics using TikZ: A Tutorial for Beginners (Part 3)---Creating Flowcharts}}. Overleaf tutorials. \url{https://www.overleaf.com/learn/latex/}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-bookdown}{}}%
Yihui Xie. 2023. \emph{{bookdown: Authoring Books and Technical Documents with R Markdown}}. \#equations. \url{https://bookdown.org/yihui/bookdown/markdown-extensions-by-bookdown.html}.

>>>>

The numbering of equations is a bit trickier. LaTeX automatically numbers equations, unless specifically instructed not to. Equation numbering in R Markdown requires specific labeling using \texttt{(\textbackslash{}\#eq:xx)} as described in Xie (2023). The \CRANpkg{texor} helps by adding the labeling using a Lua filter to convert the existing \texttt{\textbackslash{}label\{..\}} to \texttt{(\textbackslash{}\#eq:xx)}.

@@

@@

The package \texttt{texor} can be installed from CRAN and the development version from \url{https://github.com/Abhi-1U/texor}. The website for the package, \url{https://abhi-1u.github.io/texor}, has vignettes documenting usage.

@@

The package \texttt{rebib} can be installed from CRAN and the development version from \url{https://github.com/Abhi-1U/rebib}. The website for the package, \url{https://abhi-1u.github.io/rebib}, has vignettes documenting usage.

\hypertarget{the-process-of-converting-all-the-legacy-articles}{%
\section{The process of converting all the legacy articles}\label{the-process-of-converting-all-the-legacy-articles}}

\hypertarget{known-problems-needing-manual-fixes}{%
\section{Known problems needing manual fixes}\label{known-problems-needing-manual-fixes}}

@@

Since the shift to R Markdown, a new development for the open source community has emerged: quarto (Allaire et al. 2022). Although R Markdown would suggest a focus on R, it was always possible to include code chunks written in other languages. But Quarto makes this cleaner, and thus more appealing for non-R developers. It also provides a cleaner type-setting. At some point, the R Journal will likely shift to a Quarto template, which is reasonably straightforward, but for the present R Markdown is a suitable dynamic document delivery system for the R Journal.

@@

\begin{itemize}
\tightlist
\item
  \texttt{code-env}: Explains how different code environments defined by the R Journal style are handled, and additional details such as code in figure environments, and code in table environments.
\item
  \texttt{math-env}: Examples of inline math, display math, how equation numbering is handled by a Lua filter to convert from LaTeX labeling to R Markdown labeling.
\item
  \texttt{figure-env}: Explains how the variety of figure definitions are handled in the conversion, including different image formats, numbering, captions, labeling, multiple images, and \texttt{tikz} (Cassidy 2013) images.
\item
  \texttt{table-env}: Examples of how a variety of table types are converted, including multicolumn, complex and wide tables.
\item
  \texttt{lua-filters}: Overview and lots of small examples of Lua filters to handle the custom output needed for the R Markdown format.
\item
  \texttt{metadata}: This has a collection of additional format handling including extracting metadata like author names and affiliations, article identifiers used in the review process, and handling citations, footnotes and links.
\item
  \texttt{bibliography}: The bibliography was handled differently over the years of the journal, and this details how to use the \texttt{rebib} functionality to handle \texttt{bbl} files, embedded \texttt{bbl}, to convert into the standard \texttt{.bib} format.
\end{itemize}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-quarto}{}}%
Allaire, J. J., Charles Teague, Yihui Xie, and Christophe Dervieux. 2022. {``Quarto.''} Zenodo. \url{https://doi.org/10.5281/zenodo.5960048}.

\leavevmode\vadjust pre{\hypertarget{ref-tableHTML}{}}%
Boutaris, Theo, Clemens Zauchner, and Dana Jomar. 2023. \emph{tableHTML: A Tool to Create HTML Tables}. \url{https://CRAN.R-project.org/package=tableHTML}.

\leavevmode\vadjust pre{\hypertarget{ref-casflow}{}}%
Cassidy, Josh. 2013. \emph{{LaTeX Graphics using TikZ: A Tutorial for Beginners (Part 3)---Creating Flowcharts}}. Overleaf tutorials. \url{https://www.overleaf.com/learn/latex/}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-bookdown}{}}%
---------. 2023. \emph{{bookdown: Authoring Books and Technical Documents with R Markdown}}. \#equations. \url{https://bookdown.org/yihui/bookdown/markdown-extensions-by-bookdown.html}.

@@
